Aligning marine species range data to better serve science and conservation

### 2016-09-23 clip_fao/fao_clip_pac.Rmd
Added explanation for data extrapolation rate. STILL NEED TO FIX REFS TO S1 and S2 FIGURES!!!!!!!

``` {r quad_spp_w, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}

@@

# quad_names <- data.frame('quad' = c('q1', 'q2', 'q3', 'q4'),
#                          'quad_name' = factor(c('well-matched', 'dist-matched', 
#                                                 'area-matched', 'poorly matched'),
#                                              ordered = TRUE))
# 
# clipped_spp_quads <- clipped_spp_quads %>%
#   left_join(quad_names, by = 'quad') %>%
#   mutate(method = 'FAO clipped')
# 
# break_nums <- seq(0, 100, 20)

### Plot the bar chart
# barchart_clipped_spp_quads <- ggplot(clipped_spp_quads, 
#                                 aes(x = method, fill = quad, weight = pct_quad)) +
#   ggtheme_plot + 
#   theme(panel.grid.major.y = element_blank()) +
#   geom_bar(stat = 'count', alpha = 1) +
#   scale_fill_manual(values = c('q4' = '#d01c8b', 
#                                'q3' = '#f1b6da', 
#                                'q2' = '#b8e186',
#                                'q1' = '#4dac26'),
#                     labels = quad_names$quad_name) +
#   scale_y_continuous(expand = c(0, 0), 
#                      limits = c(0, 1),
#                      breaks = break_nums/100,
#                      labels = sprintf('%s%%', break_nums)) + 
#   ### add grid lines; horizontal but then get flipped
#   geom_hline(yintercept = break_nums/100, size = 0.25, color = 'white', alpha = .5) +
#   coord_flip() +
#   labs(x = '', 
#        y = 'Percent of species by quadrant', 
#        fill = 'Alignment')
# 
# print(barchart_clipped_spp_quads)

@@

``` {r combine_rasts}

stack_plot_both <- 
  tm_shape(rast_spp_w) +
    tm_raster(palette = 'Greens',
              colorNA = NULL,
              title = 'n_spp',
              alpha = 1) +
  tm_shape(rast_spp_e) +
    tm_raster(palette = 'Reds',
              colorNA = NULL,
              title = 'n_spp') +
  tm_shape(World1) +
    tm_polygons() + 
  tm_shape(fao_polys) +
    tm_borders() +
  tm_layout(basemaps = "Esri.WorldTopoMap", 
            frame = TRUE,
            # title.position = 'TOP', 
            legend.outside = TRUE, attr.outside = TRUE)

print(stack_plot_both)
```

>>>>

``` {r area_to_data_ratio, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}

@@

@@

### 2016-09-23 ms_2_methods.Rmd
Added explanation for data extrapolation rate. STILL NEED TO FIX REFS TO S1 and S2 FIGURES!!!!!!!

The IUCN has published species range maps developed by species experts for `r format(n_spp_iucn, big.mark = ',')` unique marine species.  Experts outline spatial boundaries that define the "limits of distribution" of a given species, based on observation records and informed by expert understanding of species' range and habitat preferences (Fig S1A FIX THIS REF OR THE FIGURE). IUCN guidelines generally recommend that boundaries be drawn as a "minimum convex polygon", i.e. "the smallest polygon in which no internal angle exceeds 180 degrees and which contains all the sites of occurrence" [@iucn_iucn_2015].

In contrast, AquaMaps has modeled species distribution for `r format(n_spp_am, big.mark = ',')` species based on envelopes of environmental preference, such as temperature, depth, and salinity.  These preference envelopes are deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a global 0.5° grid to determine suitable habitat, resulting in a "probability of occurrence" for each species (Fig S1B FIX THIS REF OR THE FIGURE).  To roughly constrain species ranges to appropriate georegions, the AquaMaps model uses Food and Agriculture Organization of the United Nations (FAO) Major Fishing Area [@fao_fisheries_and_aquaculture_department_cwp_2015] boundaries.  Of the resulting maps, 1,296 (5.7% of the full dataset) have been further refined through an expert review process [@kaschner_mapping_2006; @ready_predicting_2010].

@@

For each paired map species, $A_{small}$ and $A_{large}$ indicate the smaller and larger range representation (regardless of which dataset) in km^2^.  $A_{small \cap large}$ represents the amount of overlapping area between the two datasets.  Distribution alignment uses overlapping predictions of presence as a measure of concurrence between the two datasets.  Area ratio compares range size, used by IUCN as a criterion to help define extinction risk; it also provides an indicator for frequency of commission and/or omission errors.

@@

For AquaMaps data, whose reliance upon FAO Major Fishing Areas to constrain species ranges seems likely to result in abrupt discontinuities in species range where ecologically suitable habitat intersects with human-defined borders, we identified FAO boundaries where such discontinuities are likely to occur.  Cropping species distributions to narrow latitudinal bands, we determined species whose eastern or western range limit coincided exactly with a defined FAO boundary.  As an illustrative example, we cropped AquaMaps range maps for Indo-Pacific species to a band between between 25° S and 20° N latitudes, then identified all species whose eastern range limit within this band coincided with 175° W longitude, the vertical boundary between FAO regions 71 and 77.

Range maps for data-poor species are more likely to introduce errors, whether by commission (false indication of presence) or omission (false indication of absence).  This is likely to be evident as poor alignment between range maps for paired-map species.  To explore the impact of empirical measures of species occurrence on range predictions, we used R clients for OBIS [@obis_data_2016] and GBIF [@gbif_global_2010] to count occurrence records for each paired-map species; these counts were averaged for each species as a proxy for data richness.

>>>>

The IUCN has published species range maps developed by species experts for `r format(n_spp_iucn, big.mark = ',')` unique marine species.  Experts outline spatial boundaries that define the "limits of distribution" of a given species, based on observation records and informed by expert understanding of species' range and habitat preferences ___(Fig S1A FIX THIS REF OR THE FIGURE)___. IUCN guidelines recommend that boundaries be drawn as a "minimum convex polygon", i.e. "the smallest polygon in which no internal angle exceeds 180 degrees and which contains all the sites of occurrence" [@iucn_iucn_2015].

In contrast, AquaMaps has modeled species distribution for `r format(n_spp_am, big.mark = ',')` species based on envelopes of environmental preference, such as temperature, depth, and salinity.  These preference envelopes are deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a global 0.5° grid to determine suitable habitat, resulting in a "probability of occurrence" for each species ___(Fig S1B FIX THIS REF OR THE FIGURE)___.  To roughly constrain species ranges to appropriate georegions, the AquaMaps model uses Food and Agriculture Organization of the United Nations (FAO) Major Fishing Area [@fao_fisheries_and_aquaculture_department_cwp_2015] boundaries.  Of the resulting maps, 1,296 (5.7% of the full dataset) have been further refined through an expert review process [@kaschner_mapping_2006; @ready_predicting_2010].

@@

For each paired map species, $A_{small}$ and $A_{large}$ indicate the smaller and larger range representation (regardless of which dataset) in km^2^.  $A_{small \cap large}$ represents the amount of overlapping area between the two datasets.  Distribution alignment uses overlapping predictions of presence as a measure of concurrence between the two datasets.  Area ratio compares range size, used by IUCN as a criterion to help define extinction risk; it also provides an indicator for frequency of errors of commission (false indication of presence) or omission (false indication of absence).

@@

For AquaMaps data, whose reliance upon FAO Major Fishing Areas [@fao_fisheries_and_aquaculture_department_cwp_2015] to constrain species ranges seems likely to result in abrupt discontinuities in species range where ecologically suitable habitat intersects with human-defined borders, we identified FAO boundaries where such discontinuities are likely to occur.  Cropping species distributions to narrow latitudinal bands, we determined species whose eastern or western range limit coincided exactly with a defined FAO boundary.  As an illustrative example, we cropped AquaMaps range maps for all Pacific and Indian Ocean species to a band between between 25° S and 20° N latitudes, then identified all species whose eastern range limit within this band coincided with 175° W longitude, the vertical boundary between two prominent FAO regions.

### 2016-09-23 ms_3_res_disc_concl.Rmd
Added explanation for data extrapolation rate. STILL NEED TO FIX REFS TO S1 and S2 FIGURES!!!!!!!

The 3,208 species range maps included in Fig 4 tend to extrapolate farther based on limited data: a median predicted range of 452,000 km^2^ per "occurcell" compared to a median predicted range of 158,000 km^2^ per "occurcell" for the overall AquaMaps dataset.  This suggests that the FAO boundaries may not be sufficient to adequately constrain computer-generated ranges.  To reduce the incidence of commission errors due to aggressive extrapolation, it may be desirable to fine-tune the computer model output with additional filters, including ecosystem-based constraints such as Marine Ecoregions of the World [@spalding_marine_2007] or methods such as inverse distance weighting to enforce proximity to known observations.  Expert review, though time-consuming, is the most certain route to boosting confidence in these predicted distributions.

@@

To achieve more comprehensive taxonomic coverage of species ranges it may be desirable to use these two datasets together.  The Ocean Health Index (OHI) is one major analysis that merges IUCN and AquaMaps spatial data, in conjunction with extinction risk categories from the IUCN Red List of Threatened Species, to calculate area-weighted extinction risks for the Species component of the index.  To combine these data sets, OHI uses all available IUCN range maps, and then supplements with AquaMaps data (with "presence" determined by 40% or greater probability of occurrence) for species whose ranges are not represented within IUCN; species without valid Red List extinction risk categories are excluded from analysis [@halpern_index_2012; @halpern_patterns_2015].  Such a combination of two inherently different datasets increases the number of species represented within the analysis, but the differences in range representations between the datasets is likely to distort the results.

>>>>

The ratio of the total predicted range for a species to the number of "occurcells" used to generate the map provides a measure of the degree to which AquaMaps extrapolates range area from limited data.  For example, AquaMaps predicts a total range of 5.4 million km^2^ for both the round ray _Rajella fyllae_ and the brittle star _Ophiothrix plana_; but the map for _R. fyllae_ is generated using 116 "occurcells" (for a data extrapolation rate of 46,800 km^2^ per cell) while the map for _O. plana_ is generated using only four (for an extrapolation rate of 1,360,000 km^2^ per cell).

By this measure, the 3,208 species range maps included in Fig 4 tend to extrapolate farther based on limited data: a median predicted range of 452,000 km^2^ per "occurcell" compared to a median predicted range of 158,000 km^2^ per "occurcell" for the overall AquaMaps dataset.  This suggests that the FAO boundaries may not be sufficient to adequately constrain computer-generated ranges.  To reduce the incidence of commission errors due to aggressive extrapolation, it may be desirable to fine-tune the computer model output with additional filters, including ecosystem-based constraints such as Marine Ecoregions of the World [@spalding_marine_2007] or methods such as inverse distance weighting to enforce proximity to known observations.  Expert review, though time-consuming, is the most certain route to boosting confidence in these predicted distributions.

@@

To achieve more comprehensive taxonomic coverage of species ranges it may be desirable to use these two datasets together.  The Ocean Health Index (OHI) is one analysis that uses both IUCN and AquaMaps spatial data, in conjunction with extinction risk categories from the IUCN Red List of Threatened Species, to calculate area-weighted extinction risks for the Species component of the index.  To combine these data sets, OHI uses all available IUCN range maps, and then supplements with AquaMaps data (with "presence" determined by 40% or greater probability of occurrence) for species whose ranges are not represented within IUCN; species without valid Red List extinction risk categories are excluded from analysis [@halpern_index_2012; @halpern_patterns_2015].  Such a combination of two inherently different datasets increases the number of species represented within the analysis, but the differences in range representations between the datasets is likely to distort the results.
